---
output: github_document
---

<!-- README.md is generated from README.Rmd using [duckknit](https://github.com/rundel/duckknit). Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(duckknit)
knitr::opts_chunk$set(mode = "table", comment = "")

# Rewrite duckknit's ``` duckdb fences to ``` sql so GitHub applies
# SQL syntax highlighting in the rendered README.
knitr::knit_hooks$set(document = function(x) {
  gsub("(?m)^``` duckdb", "``` sql", x, perl = TRUE)
})

# Pre-start the default duckknit session with -unsigned so unsigned extensions
# can be loaded. duckknit_start_session() only accepts a db path, not extra
# CLI flags, so we launch the process directly and register it in duckknit's
# internal session table.
options(duckknit.duckdb = "/usr/local/bin/duckdb152")
local({
  ns  <- asNamespace("duckknit")
  cmd <- duckknit:::find_duckdb()
  proc <- processx::process$new(
    command = cmd,
    args = c("-unsigned", ":memory:"),
    stdin = "|", stdout = "|", stderr = "|",
    cleanup = TRUE, cleanup_tree = TRUE
  )
  unlockBinding(".sessions", ns)
  ns$.sessions[["readme"]] <- proc
  lockBinding(".sessions", ns)
  unlockBinding(".state", ns)
  ns$.state$last_session   <- "readme"
  lockBinding(".state", ns)
})
```

```{duckdb, include = FALSE}
SET allow_extensions_metadata_mismatch=true;
LOAD 'build/release/ducknng.duckdb_extension';
```

# ducknng

`ducknng` is a pure C DuckDB extension providing a binding to the [NNG](https://nng.nanomsg.org/) scalability protocols. It is inspired by [`nanonext`](https://github.com/r-lib/nanonext) and follows its ergonomic model: transport is chosen by URL scheme, aio handles wrap pending NNG operations as SQL-visible futures, and TLS material may come from filesystem paths or in-memory PEM. On top of the raw socket and transport layer, `ducknng` adds a framed RPC envelope, Arrow IPC tabular payloads, manifest-declared methods, and an HTTP/HTTPS carrier so DuckDB sessions can reach one another over `inproc://`, `ipc://`, `tcp://`, `tls+tcp://`, `ws://`, `wss://`, `http://`, or `https://`. Long-lived handles — servers, sockets, AIO futures, TLS configs, query sessions — are manually managed at the SQL surface.

It also draws on [`mangoro`](https://github.com/sounkou-bioinfo/mangoro) for the thin versioned RPC envelope design. The optional `ducknng_quack_batch` row serializer is informed by DuckDB's experimental [`quack`](https://github.com/duckdb/duckdb-quack) extension and related Quack client work such as [`@quack-protocol/sdk`](https://github.com/tobilg/quack-protocol) and [`adbc-driver-quack`](https://github.com/gizmodata/adbc-driver-quack). In ducknng today, that Quack-derived encoding is a row-payload mode inside the ducknng RPC/session protocol; it is not yet a full Quack `POST /quack` / `application/vnd.duckdb` server or client.

## Layer map

| Layer | What it does |
| --- | --- |
| **Transport** | NNG sockets and listeners. Synchronous send/recv. One-shot AIO handles. Pipe-event telemetry. |
| **Framed RPC** | Versioned envelope carrying Arrow IPC payloads or JSON control text. The query family is always on and may mutate; unary `exec` is opt-in. `http://`/`https://` mount the same methods at the URL path. |
| **Policy** | Fast C admission: mTLS, exact peer-identity allowlists, IP/CIDR allowlists, per-service and per-principal resource limits. Optional SQL authorizer at the request boundary. |
| **Codecs** | `ducknng_parse_body(...)` and `ducknng_ncurl_table(...)` parse content-type-tagged BLOBs. Built-in providers cover JSON, Arrow IPC, ducknng frames, the ducknng `ducknng_quack_batch` batch media type, and text/raw fallback. User-registered codec hooks extend the set. |

Raw protocol clients now also have a built-in `handshake` control method for capability negotiation. It advertises `supported_serialization_modes` plus `fetch_metadata` and `session_schema` capability objects covering control-plane `correlation_id` echo, stable query-session `result_handle` values, terminal `batch_index` metadata, negotiated protocol/schema versions, and effective fetch chunk counts. Today the advertised row payload modes are `arrow_ipc_stream` and `ducknng_quack_batch`: Arrow IPC remains the broad default, while `ducknng_quack_batch` is the Quack-derived DuckDB BinarySerializer batch path used for the new comparative benchmarks. The runnable wire-level example lives in `test/rpc_smoke.R`, and the contract is pinned in `docs/protocol.md`.

## Quick tour

Build the extension first (see **Development** below), then load it:

Start a TLS server. Port 0 lets the OS assign a free port; a self-signed certificate is generated in memory — no files needed.

```{duckdb collapse=TRUE, comment=""}
-- Self-signed dev cert: key and certificate live only in DuckDB memory.
SET VARIABLE tour_tls = ducknng_self_signed_tls_config('127.0.0.1', 365, 0);

SELECT ducknng_start_server(
  'tour',                              -- service name
  'tls+tcp://127.0.0.1:0',             -- TLS; port 0 = OS-assigned
  1,                                   -- REP contexts
  134217728,                           -- recv_max_bytes (128 MiB)
  300000,                              -- session_idle_ms
  getvariable('tour_tls')::UBIGINT     -- TLS config handle
);

SET VARIABLE tour_url = (
  SELECT listen FROM ducknng_list_servers() WHERE name = 'tour'
);
SELECT starts_with(
  getvariable('tour_url'),
  'tls+tcp://127.0.0.1:'
) AS tls_listen_ready;

-- Open a REQ socket and dial immediately.
SET VARIABLE req_id = (ducknng_open_socket('req')).socket_id;
SELECT (ducknng_dial_socket(
  getvariable('req_id')::UBIGINT,
  getvariable('tour_url'),
  1000, getvariable('tour_tls')::UBIGINT
)).ok AS dialed;
```

**Raw socket send/receive.** Send the manifest request frame and decode the reply envelope.

```{duckdb collapse=TRUE, comment=""}
-- Send the 22-byte manifest call frame and decode the reply.
SELECT ok, type_name, name,
       json_array_length(json_extract(payload_text::JSON, '$.methods')) AS method_count
FROM ducknng_decode_frame(
  ducknng_request_socket_raw(
    getvariable('req_id')::UBIGINT,
    from_hex('01000000000000000000000000000000000000000000'),
    1000
  )
);
```

**One-shot URL form** — same call without holding a socket handle.

```{duckdb collapse=TRUE, comment=""}
SELECT ok, type_name, name
FROM ducknng_decode_frame(
  ducknng_request_raw(
    getvariable('tour_url'),
    from_hex('01000000000000000000000000000000000000000000'),
    1000, getvariable('tour_tls')::UBIGINT
  )
);
```

**AIO: launch two requests in parallel, collect both later.** `ducknng_request_raw_aio(...)` returns immediately with an integer AIO handle. A list collect waits until any handle is ready and returns the terminal subset; this example collects each handle separately so both replies are shown.

```{duckdb collapse=TRUE, comment=""}
CREATE TEMP TABLE tour_aios AS
SELECT ducknng_request_raw_aio(
         getvariable('tour_url'),
         from_hex('01000000000000000000000000000000000000000000'),
         1000 + i::INTEGER * 0,
         getvariable('tour_tls')::UBIGINT
       ) AS aio_id
FROM range(2) t(i);

-- A one-element collect waits for that specific handle.
SELECT count(*) AS replies,
       bool_and(c.ok) AS all_ok,
       bool_and(octet_length(c.frame) > 0) AS all_have_frames
FROM tour_aios AS a,
LATERAL ducknng_aio_collect(list_value(a.aio_id), 1000) AS c;
```

**`ducknng_ncurl` — HTTP client primitive.** The framed RPC mount accepts POST only. Mount a server on an `http://` URL and POST the manifest call frame to demonstrate that the framed RPC surface is identical over HTTP.

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'tour_http', 'http://127.0.0.1:18440/_ducknng', 1, 134217728, 300000, 0::UBIGINT
);

-- The RPC mount accepts POST. POST the 22-byte manifest frame and decode the reply.
SET VARIABLE tour_http_reply = (
  SELECT body FROM ducknng_ncurl(
    'http://127.0.0.1:18440/_ducknng',
    'POST', '[{"name":"Content-Type","value":"application/vnd.ducknng.frame"}]',
    from_hex('01000000000000000000000000000000000000000000'),
    2000, 0::UBIGINT
  )
);

SELECT ok, type_name, name,
       json_array_length(json_extract(payload_text::JSON, '$.methods')) AS method_count
FROM ducknng_decode_frame(getvariable('tour_http_reply')::BLOB);
```

**Outbound HTTP profile.** Caller SQL passes a profile id and ordinary request fields. `ducknng` checks the profile scope, injects the profile auth header inside the HTTP client path, rejects caller attempts to override that header, and keeps profile introspection redacted.

```{duckdb collapse=TRUE, comment=""}
SELECT CASE WHEN ducknng_register_http_route(
  'tour_http', 'GET', '/profile-auth',
  'SELECT * FROM ducknng_http_text(
     200,
     CASE
       WHEN ducknng_http_header(''Authorization'') = ''Bearer readme-secret''
       THEN ''authorized''
       ELSE ''denied''
     END
   )'
) THEN 1 ELSE 0 END AS route_registered;

SELECT CASE WHEN ducknng_register_http_profile(
  'tour-profile', 'http', '127.0.0.1', 18440,
  '/profile-auth', 'GET', false,
  'Authorization', 'Bearer readme-secret'
) THEN 1 ELSE 0 END AS profile_registered;

SELECT ok, status, body_text
FROM ducknng_ncurl(
  'http://127.0.0.1:18440/profile-auth',
  'GET', NULL, NULL, 2000, 0::UBIGINT,
  'tour-profile'
);

SELECT profile_id, path_prefix, method, auth_header_names_json,
       position('readme-secret' IN auth_header_names_json) = 0 AS redacted
FROM ducknng_list_http_profiles()
WHERE profile_id = 'tour-profile';

SELECT CASE WHEN ducknng_drop_http_profile('tour-profile')
  THEN 1 ELSE 0 END AS profile_dropped;
```

**Query session: open, fetch Arrow IPC, parse the batch.** `ducknng_open_query` returns a `session_id` + `session_token`. `ducknng_fetch_query_table` decodes the first Arrow batch directly as rows. The raw fetch frame carries the same Arrow bytes as a BLOB for callers that need to forward or inspect the payload.

```{duckdb collapse=TRUE, comment=""}
CREATE TEMP TABLE tour_session AS
SELECT *
FROM ducknng_open_query(
  getvariable('tour_url'),
  'SELECT i, i * i AS sq FROM range(1, 6) AS t(i)',
  0::UBIGINT, 0::UBIGINT, getvariable('tour_tls')::UBIGINT
);

SET VARIABLE tour_sid   = (SELECT session_id    FROM tour_session);
SET VARIABLE tour_token = (SELECT session_token FROM tour_session);

-- Decoded rows directly:
SELECT *
FROM ducknng_fetch_query_table(
  getvariable('tour_url'),
  getvariable('tour_sid')::UBIGINT,
  getvariable('tour_token')::VARCHAR,
  0::UBIGINT, 0::UBIGINT, getvariable('tour_tls')::UBIGINT
);
```

Raw fetch gives the Arrow IPC BLOB. `ducknng_parse_body` decodes it as a table function.

```{duckdb collapse=TRUE, comment=""}
-- Re-open a fresh session to show the raw path.
CREATE TEMP TABLE tour_session2 AS
SELECT *
FROM ducknng_open_query(
  getvariable('tour_url'),
  'SELECT i, i * i AS sq FROM range(1, 6) AS t(i)',
  0::UBIGINT, 0::UBIGINT, getvariable('tour_tls')::UBIGINT
);

SET VARIABLE tour_raw_frame = ducknng_fetch_query_raw(
  getvariable('tour_url'),
  (SELECT session_id    FROM tour_session2)::UBIGINT,
  (SELECT session_token FROM tour_session2)::VARCHAR,
  0::UBIGINT, 0::UBIGINT, getvariable('tour_tls')::UBIGINT
);

-- Inspect envelope fields, then decode the Arrow payload.
SELECT ducknng_frame_type_name(getvariable('tour_raw_frame')::BLOB) AS type_name,
       ducknng_frame_end_of_stream(getvariable('tour_raw_frame')::BLOB) AS end_of_stream;

SELECT *
FROM ducknng_parse_body(
  ducknng_frame_payload(getvariable('tour_raw_frame')::BLOB),
  'application/vnd.apache.arrow.stream'
);
```

**SSE streaming route.** `ducknng_add_stream_route` registers a chunked-streaming route. The SQL must return a `chunk` column; each non-null row is written to the client as a separate HTTP chunk. `ducknng_format_sse` formats a Server-Sent Events event string.

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_add_stream_route(
  'tour_http', 'GET', '/events',
  'SELECT ducknng_format_sse(''tick '' || i::VARCHAR) AS chunk
   FROM generate_series(1, 3) t(i)'
);

-- ncurl transparently reassembles the chunked body.
SELECT ok, status, body_text
FROM ducknng_ncurl('http://127.0.0.1:18440/events', 'GET', NULL, NULL, 2000, 0::UBIGINT);
```

Explicit cleanup — everything that was opened must be explicitly stopped, closed, or dropped.

```{duckdb collapse=TRUE, comment=""}
SELECT bool_and(ducknng_aio_drop(aio_id)) AS aios_dropped
FROM tour_aios;
DROP TABLE tour_aios;
DROP TABLE tour_session;
DROP TABLE tour_session2;
SELECT (ducknng_close_socket(getvariable('req_id')::UBIGINT)).ok AS socket_closed;
SELECT ducknng_stop_server('tour_http');
SELECT ducknng_stop_server('tour');
SELECT ducknng_drop_tls_config(getvariable('tour_tls')::UBIGINT);
```

## Development

Prerequisites: a C compiler, CMake, Python 3, and R with the `rmarkdown` and [`duckknit`](https://github.com/rundel/duckknit) packages.

```sh
# Download DuckDB headers and set up the build tree. Run once per checkout.
make configure

# Build the release extension binary.
# Output: build/release/ducknng.duckdb_extension
make release

# Run the SQL test suite against the release build.
make test_release

# Run the ducknng-only RPC microbench harness.
make rpc_bench

# Render the bulk benchmark report to bench/rpc_bulk_compare.md.
# This compares ducknng RPC over http/tcp/ipc/ws in both arrow_ipc_stream
# and ducknng_quack_batch modes, plus Quack over its public client path.
# Requires network access to INSTALL quack FROM core_nightly.
make rpc_bulk_compare

# Render README.md and demo/subscriber_gateway.md from their Rmd sources.
make docs
```

`make configure` downloads the DuckDB C API headers at the version pinned in the Makefile (`DUCKDB_HEADER_VERSION`) and sets up CMake. The extension targets `TARGET_DUCKDB_VERSION`. It builds with `USE_UNSTABLE_C_API=1` to access DuckDB's native Arrow conversion API (`duckdb_to_arrow_schema`, `duckdb_data_chunk_to_arrow`) in the result-to-IPC emit path. The only deprecated-entrypoint exception is the pending-result streaming pair, isolated in `src/ducknng_duckdb_streaming_compat.c`, because DuckDB v1.5.2 does not yet expose an undeprecated C API for pending execution with incremental chunk delivery. Query sessions do not silently fall back to materialized results; when DuckDB provides a replacement streaming API, that compatibility boundary is the single place to change. `make rpc_bench` keeps the existing raw-RPC microbench path. `make rpc_bulk_compare` now renders `bench/rpc_bulk_compare.md`, a benchmark report that records machine details, compares ducknng RPC over `http`, `tcp`, `ipc`, and `ws` in both `arrow_ipc_stream` and `ducknng_quack_batch` modes, compares Quack over its public client/server path, and includes concurrent reader, writer, and mixed reader/writer RPC slices. The writer slice is a set-oriented remote SQL write dispatch benchmark, not a client-side bulk append protocol. The report generates a local TPC-H dataset if needed and installs `quack` from `core_nightly`. The unstable functions in use are tracked by `tools/used_duckdb_unstable_api.R`; the table below is generated fresh on every README render:

```{r unstable-api-table, echo = FALSE, results = "asis"}
source("tools/used_duckdb_unstable_api.R")
cat(ducknng_used_duckdb_unstable_api_markdown())
```

`make test_release` runs the SQL test suite with DuckDB's test runner. All tests live under `test/sql/` as `.test` files. The public function surface is verified by `test/sql/ducknng_public_surface.test`, which reads `function_catalog/functions.yaml` and checks that every declared name appears in the loaded extension catalog.

To regenerate the function catalog markdown (embedded in this README) after adding or renaming public functions:

```sh
make function_catalog  # regenerates function_catalog/functions.md
```

Every new public SQL function must have a corresponding entry in `function_catalog/functions.yaml` or the surface test will fail.

## Lifetime and manual cleanup

Long-lived handles are manually managed at the SQL surface. Stop, close, or drop them explicitly; runtime teardown is fallback cleanup, not the primary lifecycle model.

- Stop servers with `ducknng_stop_server(...)`
- Close sockets with `ducknng_close_socket(...)`
- Drop AIO handles with `ducknng_aio_drop(...)`
- Drop TLS config handles with `ducknng_drop_tls_config(...)`
- Close query sessions with `ducknng_close_query(...)`

`ducknng_aio_cancel(...)` is cancellation control, not the destructor. `ducknng_cancel_query(...)` is best-effort session control, not a replacement for the explicit close path.

## Function catalog

The generated catalog is the public SQL surface. Any loaded `ducknng__*` helper names are internal macro implementation details required by DuckDB's stable C API path and are not supported user APIs.

```{r function-catalog, echo=FALSE, results='asis'}
catalog <- paste(readLines("function_catalog/functions.md"), collapse = "\n")
knitr::asis_output(paste0(
  "<details>\n",
  "<summary><strong>Expand the full generated function catalog</strong></summary>\n\n",
  catalog,
  "\n\n</details>\n"
))
```

## Transport layer

### NNG socket patterns

NNG exposes seven socket patterns. All share the same `ducknng_open_socket(kind)` / `ducknng_listen_socket(...)` / `ducknng_dial_socket(...)` entry points and the same synchronous/AIO send-recv helpers.

| Pattern | `kind` | Direction | Typical use |
| --- | --- | --- | --- |
| Request/reply | `req` / `rep` | 1:1 round-trip | RPC, sync calls |
| Pipeline | `push` / `pull` | 1-way fan-out | Task queues, load distribution |
| Publish/subscribe | `pub` / `sub` | 1:N broadcast | Event fan-out; subscribers filter by prefix |
| Survey | `surveyor` / `respondent` | 1:N with replies | Health checks, membership queries |
| Bus | `bus` | N:N mesh | Fully-connected peer broadcast |
| Pair | `pair` | 1:1 bidirectional | Point-to-point channels |

The push/pull example below shows the pattern shared by all of them: arm an async receiver, send synchronously, collect the AIO result.

```{duckdb collapse=TRUE, comment=""}
SET VARIABLE pull_s = (ducknng_open_socket('pull')).socket_id;
SET VARIABLE pull_ok = (ducknng_listen_socket(
  getvariable('pull_s')::UBIGINT,
  'ipc:///tmp/ducknng_readme_pushpull.ipc',
  134217728, 0::UBIGINT
)).ok;
SET VARIABLE push_s = (ducknng_open_socket('push')).socket_id;
SET VARIABLE push_ok = (ducknng_dial_socket(
  getvariable('push_s')::UBIGINT,
  'ipc:///tmp/ducknng_readme_pushpull.ipc',
  1000, 0::UBIGINT
)).ok;

CREATE TEMP TABLE pp_recv AS
SELECT ducknng_recv_socket_raw_aio(getvariable('pull_s')::UBIGINT, 1000) AS recv_aio;

SET VARIABLE pp_sent = (ducknng_send_socket_raw(
  getvariable('push_s')::UBIGINT, from_hex('6865796f'), 1000
)).ok;

SELECT getvariable('pp_sent')::BOOLEAN AS sent,
       ok, hex(frame) = '6865796F' AS got_payload
FROM ducknng_aio_collect((SELECT list_value(recv_aio) FROM pp_recv), 1000);

SELECT ducknng_aio_drop((SELECT recv_aio FROM pp_recv)) AS dropped;
DROP TABLE pp_recv;
SELECT (ducknng_close_socket(getvariable('push_s')::UBIGINT)).ok AS closed_push,
       (ducknng_close_socket(getvariable('pull_s')::UBIGINT)).ok AS closed_pull;
```

For pub/sub, call `ducknng_subscribe_socket(socket_id, prefix_blob)` before dialing. For surveyor, send then arm multiple respondent receivers. For bus, every peer dials the same listener so each send reaches all other peers.

### TLS (`tls+tcp://`, `wss://`, `https://`)

TLS handles are created once and reused across services and clients. Material can come from the filesystem or from in-memory PEM content.

```{duckdb collapse=TRUE, comment=""}
-- Self-signed development cert: no files needed.
SET VARIABLE tls_self = ducknng_self_signed_tls_config('127.0.0.1', 365, 0);

SELECT ducknng_start_server(
  'tls_demo', 'tls+tcp://127.0.0.1:45453',
  1, 134217728, 300000,
  getvariable('tls_self')::UBIGINT
);

SELECT ok, type_name, name
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'tls+tcp://127.0.0.1:45453',
    from_hex('01000000000000000000000000000000000000000000'),
    1000,
    getvariable('tls_self')::UBIGINT
  )
);

SELECT ducknng_stop_server('tls_demo');
SELECT ducknng_drop_tls_config(getvariable('tls_self')::UBIGINT);
```

File-backed material uses `ducknng_tls_config_from_files(cert_key_file, ca_file, password, auth_mode)`. TLS client handles verify the remote server certificate by default when `auth_mode = 0`; on listeners, `auth_mode = 0` means ordinary server TLS without requiring client certificates. Set listener `auth_mode = 2` to require client certificates (mTLS). The dispatcher derives caller identity from the first verified SAN (`tls:san:<value>`) falling back to CN (`tls:cn:<name>`). Peer identity and IP/CIDR allowlists narrow admission before SQL dispatch.

```{duckdb collapse=TRUE, comment=""}
SET VARIABLE tls_files = ducknng_tls_config_from_files(
  'test/certs/loopback-cert-key.pem',
  'test/certs/loopback-ca.pem',
  NULL, 0
);

SELECT ducknng_start_server(
  'tls_files_demo', 'tls+tcp://127.0.0.1:45454',
  1, 134217728, 300000,
  getvariable('tls_files')::UBIGINT
);

SELECT ok, type_name, name
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'tls+tcp://127.0.0.1:45454',
    from_hex('01000000000000000000000000000000000000000000'),
    1000,
    getvariable('tls_files')::UBIGINT
  )
);

SELECT ducknng_stop_server('tls_files_demo');
SELECT ducknng_drop_tls_config(getvariable('tls_files')::UBIGINT);
```

### WebSocket (`ws://`, `wss://`)

`ws://` and `wss://` are NNG transport schemes. They use `ducknng_start_server(...)`, the full framed RPC surface, and the same TLS handle model as `tls+tcp://`. They are not the HTTP carrier described in `docs/http.md`.

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'ws_demo', 'ws://127.0.0.1:45455/_ducknng', 1, 134217728, 300000, 0::UBIGINT
);

SELECT ok, type_name, name
FROM ducknng_decode_frame(
  ducknng_request_raw(
    'ws://127.0.0.1:45455/_ducknng',
    from_hex('01000000000000000000000000000000000000000000'),
    1000, 0::UBIGINT
  )
);

SELECT ducknng_stop_server('ws_demo');
```

## Framed RPC

### Registry: manifest and exec

Every server starts with a `manifest` method and nothing else. `exec` is registered explicitly.

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'rpc_demo', 'ipc:///tmp/ducknng_readme_rpc.ipc', 1, 134217728, 300000, 0
);

-- Default registry: manifest only.
SELECT name, family, response_mode, requires_auth, disabled
FROM ducknng_list_methods()
ORDER BY name;

-- Register exec (false = no auth required).
SELECT ducknng_register_exec_method(false) AS registered_exec;

-- Manifest now reports two methods.
WITH m AS (
  SELECT manifest
  FROM ducknng_get_rpc_manifest('ipc:///tmp/ducknng_readme_rpc.ipc', 0::UBIGINT)
  WHERE ok
)
SELECT json_extract_string(manifest::JSON, '$.server.name') AS server_name,
       json_array_length(json_extract(manifest::JSON, '$.methods')) AS method_count,
       position('"name":"exec"' IN manifest) > 0 AS has_exec
FROM m;
```

### Registry: the upload lane

The upload lane is the client-to-server mirror of the query family: instead of
fetching result batches, a client streams `ducknng_quack_batch` tuples *into* a
server table. Like `exec`, it is opt-in and exposes no surface until the host
registers it. `ducknng_register_upload_methods(false)` registers it without an
auth requirement (the no-argument form defaults the same way); it reuses the
query family's execution-subject admission, session owner tokens, and
per-peer-identity limits.

```{duckdb collapse=TRUE, comment=""}
-- Register the upload lane (false = no auth required), mirroring exec.
SELECT ducknng_register_upload_methods(false) AS registered_upload;

-- The four upload-family methods are now advertised.
SELECT name, family, response_mode, requires_auth, disabled
FROM ducknng_list_methods()
WHERE family = 'upload'
ORDER BY name;
```

The flow is `upload_open` → `upload_append`* → `upload_commit` (or
`upload_abort`). `upload_open` takes JSON `{target_table, mode}` (v1 appends to
an existing table), acquires a session connection, begins a transaction, and
opens a DuckDB appender. Each `upload_append` body is a counted control prefix
(`[session_id u64 LE][token_len u16 LE][token]`) followed by a raw quack batch,
which the server decodes straight into the session appender; the batch's column
names, order, and types must match the target table exactly, since the appender
appends by ordinal. `upload_commit` flushes and commits; `upload_abort` — and
any idle prune or shutdown — rolls the transaction back so partially uploaded
rows never persist. The counted token is matched over its full length, and
upload and query control methods reject each other's sessions. The wire contract
is pinned in `docs/protocol.md`. The ergonomic
`ducknng_upload_table(url, source_query, target_table[, tls_config_id])` table
function drives this end-to-end: it runs `source_query` locally and streams the
result rows into `target_table` over the upload lane (the batch column names,
order, and types must match the target), returning the uploaded row and
client-sent byte counts. A runnable round-trip is in
`test/sql/ducknng_upload_roundtrip.test`.

### Synchronous helpers

```{duckdb collapse=TRUE, comment=""}
-- DDL statement over RPC.
SELECT * FROM ducknng_run_rpc(
  'ipc:///tmp/ducknng_readme_rpc.ipc',
  'CREATE TABLE rpc_demo_t(i INTEGER)',
  0::UBIGINT
);

-- DML over RPC.
SELECT * FROM ducknng_run_rpc(
  'ipc:///tmp/ducknng_readme_rpc.ipc',
  'INSERT INTO rpc_demo_t VALUES (10),(11),(12)',
  0::UBIGINT
);

-- Query rows over RPC — decoded as a table.
SELECT *
FROM ducknng_query_rpc(
  'ipc:///tmp/ducknng_readme_rpc.ipc',
  'SELECT i, i > 10 AS gt_10 FROM rpc_demo_t ORDER BY i',
  0::UBIGINT
);
```

The `_params` helpers carry a typed Arrow `STRUCT`; child order binds SQL `?`
parameters and values are never interpolated into SQL text. Use explicit SQL
type context for a null parameter, such as `?::BIGINT`.

```{duckdb collapse=TRUE, comment=""}
SELECT rows_changed
FROM ducknng_run_rpc_params(
  'ipc:///tmp/ducknng_readme_rpc.ipc',
  'INSERT INTO rpc_demo_t VALUES (?)',
  struct_pack(value := 13::INTEGER),
  0::UBIGINT
);

SELECT i
FROM ducknng_query_rpc_params(
  'ipc:///tmp/ducknng_readme_rpc.ipc',
  'SELECT i FROM rpc_demo_t WHERE i >= ? ORDER BY i',
  struct_pack(minimum := 11::INTEGER),
  0::UBIGINT
);

DESCRIBE SELECT *
FROM ducknng_prepare_query_params(
  'ipc:///tmp/ducknng_readme_rpc.ipc',
  'SELECT i FROM rpc_demo_t WHERE i >= ?',
  struct_pack(minimum := 11::INTEGER),
  0::UBIGINT
);
```

The query helpers discover the remote result schema during DuckDB bind, which
may happen more than once; use them for read-only or idempotent SQL. The
fixed-schema `run_rpc` helpers send `exec` during scan, so `EXPLAIN` does not
perform the remote statement. This avoids bind-time mutation but does not make
a network write exactly once: after a lost reply, retry policy belongs to the
application.

You can also request the Quack-derived batch serializer over the same session contract. Unlike upstream [Quack](https://github.com/duckdb/duckdb-quack), ducknng keeps transport choice separate from row serialization, so the same `ducknng_quack_batch` mode can ride over `http`, `tcp`, `ipc`, or `ws`. Version and schema compatibility belong to the negotiated ducknng session contract rather than to a suffix on the serializer token; full Quack protocol traffic uses `application/vnd.duckdb` messages and negotiates `quackVersion` during the connection exchange, but this fetch payload is not wrapped in that Quack connection envelope. `ducknng.fetch_batch_chunks` controls the default number of DuckDB chunks requested per fetch; it defaults to 12 and applies to Arrow IPC and `ducknng_quack_batch` alike unless the client sends an explicit `batch_rows` hint.

```{duckdb collapse=TRUE, comment=""}
SELECT *
FROM ducknng_query_rpc_mode(
  'ipc:///tmp/ducknng_readme_rpc.ipc',
  'SELECT i, i > 10 AS gt_10 FROM rpc_demo_t ORDER BY i',
  0::UBIGINT,
  'ducknng_quack_batch'
);
```

The Quack-derived mode also carries nested DuckDB values through the same RPC/session lifecycle. Its decoder materializes DuckDB `CONSTANT`, `DICTIONARY`, and integer `SEQUENCE` vectors into flat C API output, including recursive dictionary values, while rejecting malformed selections, unbounded compressed expansion, and unsupported FSST vectors. Canonical chunks repeat and validate the `DataChunk` logical-type list. The pure-C encoder remains flat because DuckDB's C API does not expose source compression state. A normal SQL query cannot reliably force a particular compressed source-vector shape through that API, so the executable compressed-vector fixtures are generated by `tools/quack_compressed_fixtures.R` and asserted in `test/sql/ducknng_body_codecs.test` rather than rendered as opaque hexadecimal output here.

This is not a claim over every logical type in every DuckDB release. DuckDB v1.5.2 has `GEOMETRY` and `VARIANT`, but the current Quack-derived codec rejects both: geometry carries versioned format and optional CRS metadata, while that release's C API has no `DUCKDB_TYPE_VARIANT` constructor or vector contract. They are not silently projected to `BLOB` or `STRUCT`.

```{duckdb collapse=TRUE, comment=""}
SELECT st.a = 7 AS struct_ok,
       st.xs = [1, 2, 3]::INTEGER[] AS nested_list_ok
FROM ducknng_query_rpc_mode(
  'ipc:///tmp/ducknng_readme_rpc.ipc',
  'SELECT {''a'': 7::INTEGER, ''xs'': [1,2,3]::INTEGER[]} AS st',
  0::UBIGINT,
  'ducknng_quack_batch'
);
```

Arrow IPC carries temporal, decimal, list, struct, union, large-string, large-binary, fixed-size-binary, and duration values through the same path. Supported mappings are documented in `docs/types.md`.

```{duckdb collapse=TRUE, comment=""}
SELECT d = DATE '2024-01-02' AS date_ok,
       ts = TIMESTAMP '2024-01-02 03:04:05.123456' AS ts_ok,
       dec = 123.45::DECIMAL(10,2) AS decimal_ok,
       xs[2] IS NULL AND xs[3] = 3 AS list_ok,
       st.a = 7 AND st.b = 'bee' AS struct_ok
FROM ducknng_query_rpc(
  'ipc:///tmp/ducknng_readme_rpc.ipc',
  'SELECT DATE ''2024-01-02'' AS d,
          TIMESTAMP ''2024-01-02 03:04:05.123456'' AS ts,
          123.45::DECIMAL(10,2) AS dec,
          [1, NULL, 3]::INTEGER[] AS xs,
          {''a'': 7::INTEGER, ''b'': ''bee''} AS st',
  0::UBIGINT
);
```

`ducknng_parse_body(...)` decodes content-type-tagged BLOBs. The built-in providers cover Arrow IPC, JSON, ducknng frames, the ducknng `application/vnd.ducknng.quack-batch` batch media type, and a text/raw fallback.

```{duckdb collapse=TRUE, comment=""}
SELECT provider, output FROM ducknng_list_codecs() ORDER BY provider;

SELECT a, b
FROM ducknng_parse_body(
  '[{"a":1,"b":"x"},{"a":2,"b":"y"}]'::BLOB,
  'application/json; charset=utf-8'
)
ORDER BY a;
```

### Raw frame helpers

The raw scalar forms return reply bytes as BLOBs. `ducknng_decode_frame(...)` projects all envelope fields, including the protocol status, plus the text payload. `ducknng_frame_status(...)` returns that status directly; it is distinct from an NNG transport error code.

```{duckdb collapse=TRUE, comment=""}
-- Raw bytes from a socket-scoped request (show first 14 bytes as hex).
SET VARIABLE req2 = (ducknng_open_socket('req')).socket_id;
SET VARIABLE dialed2 = (ducknng_dial_socket(
  getvariable('req2')::UBIGINT,
  'ipc:///tmp/ducknng_readme_rpc.ipc',
  1000, 0::UBIGINT
)).ok;

SELECT substr(
  hex(ducknng_request_socket_raw(
    getvariable('req2')::UBIGINT,
    from_hex('01000000000000000000000000000000000000000000'),
    1000
  )),
  1, 28
) AS first_14_bytes_hex;

-- Decode a manifest reply directly.
SELECT ok, version, type_name, status, name,
       position('"name":"exec"' IN payload_text) > 0 AS has_exec
FROM ducknng_decode_frame(
  ducknng_get_rpc_manifest_raw('ipc:///tmp/ducknng_readme_rpc.ipc', 0::UBIGINT)
);

-- A malformed request returns an INVALID protocol status in its error frame.
SELECT ducknng_frame_status(ducknng_request_raw(
  'ipc:///tmp/ducknng_readme_rpc.ipc', from_hex('00'), 1000, 0::UBIGINT
)) AS invalid_status;

SELECT (ducknng_close_socket(getvariable('req2')::UBIGINT)).ok AS closed;
```

### AIO helpers

`ducknng_request_raw_aio(...)` and `ducknng_get_rpc_manifest_raw_aio(...)` return integer AIO handles immediately. `ducknng_aio_collect(...)` blocks until all handles are ready and returns one decoded-frame row per handle. `ducknng_aio_wait(...)` waits without consuming — the handle stays inspectable and droppable afterward. `ducknng_aio_collect_decoded(...)` projects envelope columns directly.

```{duckdb collapse=TRUE, comment=""}
-- Launch manifest and exec requests in parallel.
SET VARIABLE aio_manifest = ducknng_get_rpc_manifest_raw_aio(
  'ipc:///tmp/ducknng_readme_rpc.ipc', 1000, 0::UBIGINT
);
SET VARIABLE aio_exec = ducknng_run_rpc_raw_aio(
  'ipc:///tmp/ducknng_readme_rpc.ipc',
  'CREATE TABLE IF NOT EXISTS rpc_aio_t(i INTEGER)',
  1000, 0::UBIGINT
);

SELECT getvariable('aio_manifest') > 0 AS manifest_launched,
       getvariable('aio_exec') > 0    AS exec_launched;

-- Collect both frames when ready.
CREATE TEMP TABLE aio_results AS
SELECT * FROM ducknng_aio_collect(
  list_value(getvariable('aio_manifest'), getvariable('aio_exec')), 1000
);

SET VARIABLE frame_manifest = (SELECT frame FROM aio_results WHERE aio_id = getvariable('aio_manifest'));
SET VARIABLE frame_exec     = (SELECT frame FROM aio_results WHERE aio_id = getvariable('aio_exec'));

SELECT ok, type_name, name,
       position('"name":"exec"' IN payload_text) > 0 AS has_exec
FROM ducknng_decode_frame(getvariable('frame_manifest'));

SELECT ok, type_name, name
FROM ducknng_decode_frame(getvariable('frame_exec'));

-- collect_decoded projects envelope columns directly without a second decode call.
SET VARIABLE aio_manifest2 = ducknng_get_rpc_manifest_raw_aio(
  'ipc:///tmp/ducknng_readme_rpc.ipc', 1000, 0::UBIGINT
);
SELECT aio_id, ok, frame_ok, type_name, name
FROM ducknng_aio_collect_decoded(list_value(getvariable('aio_manifest2')), 1000);

SELECT ducknng_aio_drop(getvariable('aio_manifest')) AND
       ducknng_aio_drop(getvariable('aio_exec')) AND
       ducknng_aio_drop(getvariable('aio_manifest2')) AS dropped;
DROP TABLE aio_results;
SELECT ducknng_stop_server('rpc_demo');
```

## Query sessions

Query sessions decouple opening a query from fetching its result batches. The server holds the cursor; the client drives the fetch loop using `session_id` + `session_token`. The token is the bearer capability that authorizes fetch, close, and cancel.

### High-level (open / fetch_table / close)

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'session_demo', 'ipc:///tmp/ducknng_readme_session.ipc', 1, 134217728, 300000, 0
);

CREATE TEMP TABLE sess AS
SELECT *
FROM ducknng_open_query(
  'ipc:///tmp/ducknng_readme_session.ipc',
  'SELECT 1 AS id UNION ALL SELECT 2 AS id ORDER BY id',
  0::UBIGINT, 0::UBIGINT, 0::UBIGINT
);

SET VARIABLE sid   = (SELECT session_id    FROM sess);
SET VARIABLE stok  = (SELECT session_token FROM sess);

-- Decode first batch as rows directly.
SELECT * FROM ducknng_fetch_query_table(
  'ipc:///tmp/ducknng_readme_session.ipc',
  getvariable('sid')::UBIGINT,
  getvariable('stok')::VARCHAR,
  0::UBIGINT, 0::UBIGINT, 0::UBIGINT
);

-- Next fetch returns the exhausted control reply.
SELECT ok, state, end_of_stream
FROM ducknng_fetch_query(
  'ipc:///tmp/ducknng_readme_session.ipc',
  getvariable('sid')::UBIGINT,
  getvariable('stok')::VARCHAR,
  0::UBIGINT, 0::UBIGINT, 0::UBIGINT
);

SELECT ok, state
FROM ducknng_close_query(
  'ipc:///tmp/ducknng_readme_session.ipc',
  getvariable('sid')::UBIGINT,
  getvariable('stok')::VARCHAR,
  0::UBIGINT
);

DROP TABLE sess;
SELECT ducknng_stop_server('session_demo');
```

### Raw frames (open_raw / fetch_raw / close_raw)

The raw helpers return the reply envelope as a BLOB. `ducknng_frame_payload(...)` extracts the Arrow IPC bytes; `ducknng_parse_body(...)` turns them into rows.

```{duckdb collapse=TRUE, comment=""}
SET VARIABLE session_raw_tls = ducknng_self_signed_tls_config('127.0.0.1', 365, 0);

SELECT ducknng_start_server(
  'session_raw', 'tls+tcp://127.0.0.1:0', 1, 134217728, 300000,
  getvariable('session_raw_tls')::UBIGINT
);

SET VARIABLE session_raw_url = (
  SELECT listen FROM ducknng_list_servers() WHERE name = 'session_raw'
);

SET VARIABLE raw_open = ducknng_open_query_raw(
  getvariable('session_raw_url'),
  'SELECT i, i * i AS sq FROM range(1, 4) AS t(i)',
  0::UBIGINT, 0::UBIGINT, getvariable('session_raw_tls')::UBIGINT
);

SET VARIABLE raw_sid = (
  SELECT json_extract(
    ducknng_frame_payload_text(getvariable('raw_open')::BLOB)::JSON,
    '$.session_id'
  )::UBIGINT
);
SET VARIABLE raw_tok = (
  SELECT json_extract_string(
    ducknng_frame_payload_text(getvariable('raw_open')::BLOB)::JSON,
    '$.session_token'
  )
);

SET VARIABLE raw_fetch = ducknng_fetch_query_raw(
  getvariable('session_raw_url'),
  getvariable('raw_sid')::UBIGINT,
  getvariable('raw_tok')::VARCHAR,
  0::UBIGINT, 0::UBIGINT, getvariable('session_raw_tls')::UBIGINT
);

SELECT ducknng_frame_type_name(getvariable('raw_fetch')::BLOB)      AS type_name,
       ducknng_frame_end_of_stream(getvariable('raw_fetch')::BLOB)  AS end_of_stream;

SELECT *
FROM ducknng_parse_body(
  ducknng_frame_payload(getvariable('raw_fetch')::BLOB),
  'application/vnd.apache.arrow.stream'
);

SET VARIABLE raw_close = ducknng_close_query_raw(
  getvariable('session_raw_url'),
  getvariable('raw_sid')::UBIGINT,
  getvariable('raw_tok')::VARCHAR,
  getvariable('session_raw_tls')::UBIGINT
);

SELECT position('"state":"closed"' IN
  ducknng_frame_payload_text(getvariable('raw_close')::BLOB)
) > 0 AS is_closed;

SELECT ducknng_stop_server('session_raw');
SELECT ducknng_drop_tls_config(getvariable('session_raw_tls')::UBIGINT);
```

### Async (open_aio / fetch_aio / close_aio / cancel_aio)

Each session lifecycle step has a `_raw_aio` variant. The AIO handle is collected through `ducknng_aio_collect(...)` just like any other AIO handle.

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'session_aio', 'ipc:///tmp/ducknng_readme_session_aio.ipc', 1, 134217728, 300000, 0
);

SET VARIABLE open_aio = ducknng_open_query_raw_aio(
  'ipc:///tmp/ducknng_readme_session_aio.ipc',
  'SELECT 99 AS id',
  0::UBIGINT, 0::UBIGINT, 1000, 0::UBIGINT
);

CREATE TEMP TABLE open_collect AS
SELECT * FROM ducknng_aio_collect(list_value(getvariable('open_aio')::UBIGINT), 1000);

SET VARIABLE open_frame = (SELECT frame FROM open_collect);
SET VARIABLE aio_sid = (
  SELECT json_extract(payload_text::JSON, '$.session_id')::UBIGINT
  FROM ducknng_decode_frame(getvariable('open_frame'))
);
SET VARIABLE aio_tok = (
  SELECT json_extract_string(payload_text::JSON, '$.session_token')
  FROM ducknng_decode_frame(getvariable('open_frame'))
);

SET VARIABLE close_aio = ducknng_close_query_raw_aio(
  'ipc:///tmp/ducknng_readme_session_aio.ipc',
  getvariable('aio_sid')::UBIGINT,
  getvariable('aio_tok')::VARCHAR,
  1000, 0::UBIGINT
);

CREATE TEMP TABLE close_collect AS
SELECT * FROM ducknng_aio_collect(list_value(getvariable('close_aio')::UBIGINT), 1000);

SET VARIABLE close_frame = (SELECT frame FROM close_collect);
SELECT ok, type_name, name,
       position('"state":"closed"' IN payload_text) > 0 AS is_closed
FROM ducknng_decode_frame(getvariable('close_frame'));

SELECT ducknng_aio_drop(getvariable('open_aio')::UBIGINT) AND
       ducknng_aio_drop(getvariable('close_aio')::UBIGINT) AS dropped;
DROP TABLE open_collect; DROP TABLE close_collect;
SELECT ducknng_stop_server('session_aio');
```

## HTTP carrier

### `ducknng_ncurl` — HTTP client primitive

`ducknng_ncurl(url, method, headers_json, body, timeout_ms, tls_config_id)` is the low-level HTTP/HTTPS transport primitive. Its `headers_json` argument uses `[{"name":"...","value":"..."}]` arrays; header names must be HTTP tokens and values must not contain CRLF or other control characters. A local `nanonext` HTTPS server is started during README rendering for the TLS example; the server definition is shown because it is part of the example rather than hidden setup.

```{r http-demo-setup, echo=FALSE, results='asis'}
library(nanonext)
pidfile <- "/tmp/ducknng_readme_http_demo.pid"
script  <- "/tmp/ducknng_readme_http_demo.R"
logfile <- "/tmp/ducknng_readme_http_demo.log"
cafile  <- "/tmp/ducknng_readme_http_demo_ca.pem"
if (file.exists(pidfile)) {
  pid <- suppressWarnings(as.integer(readLines(pidfile, warn = FALSE)))
  if (!is.na(pid)) tools::pskill(pid)
  unlink(pidfile)
}
http_demo_code <- paste(
  "library(nanonext)",
  "",
  "cert <- write_cert(cn = '127.0.0.1')",
  sprintf("writeLines(cert$client[[1]], %s)", deparse(cafile)),
  "",
  "server <- http_server(",
  "  url = 'https://127.0.0.1:18443',",
  "  handlers = list(",
  "    handler('/hello', function(req) {",
  "      list(status = 200L,",
  "           headers = c('Content-Type' = 'text/plain', 'X-Test' = 'hello'),",
  "           body = 'hello from nanonext https server')",
  "    }),",
  "    handler('/echo', function(req) {",
  "      list(status = 200L,",
  "           headers = c('Content-Type' = (",
  "             req$headers[['Content-Type']] %||% 'application/octet-stream'),",
  "             'X-Test' = 'echo'),",
  "           body = req$body)",
  "    }, method = 'POST')",
  "  ),",
  "  tls = tls_config(server = cert$server)",
  ")",
  sep = "\n"
)
cat("```r\n", http_demo_code, "\n```\n\n", sep = "")
launch_code <- paste(
  http_demo_code, "",
  "library(later)", "attr(server, 'start')()", "repeat later::run_now(Inf)",
  sep = "\n"
)
writeLines(launch_code, script)
status <- system(sprintf(
  "Rscript %s > %s 2>&1 & echo $! > %s",
  shQuote(script), shQuote(logfile), shQuote(pidfile)
))
if (status != 0 || !file.exists(pidfile))
  stop("failed to start the README HTTP demo server")
pid <- suppressWarnings(as.integer(readLines(pidfile, warn = FALSE)))
if (is.na(pid) || system(sprintf("kill -0 %d >/dev/null 2>&1", pid)) != 0) {
  log_text <- if (file.exists(logfile)) paste(readLines(logfile, warn = FALSE), collapse = "\n") else ""
  stop("README HTTP demo server did not stay running\n", log_text)
}
ready <- FALSE
for (attempt in seq_len(20)) {
  client_tls <- tryCatch(
    if (file.exists(cafile)) tls_config(client = cafile) else NULL,
    error = function(e) NULL
  )
  probe <- tryCatch(
    if (!is.null(client_tls))
      ncurl("https://127.0.0.1:18443/hello", timeout = 500L, tls = client_tls)
    else NULL,
    error = function(e) NULL
  )
  if (!is.null(probe) && identical(probe$status, 200L)) { ready <- TRUE; break }
  Sys.sleep(0.25)
}
if (!ready) {
  log_text <- if (file.exists(logfile)) paste(readLines(logfile, warn = FALSE), collapse = "\n") else ""
  stop("README HTTP demo server did not become ready\n", log_text)
}
```

```{duckdb collapse=TRUE, comment=""}
-- Register a client TLS handle that trusts the self-signed demo CA.
-- In client mode auth_mode = 0 still verifies the HTTPS server certificate.
SET VARIABLE tls_http = ducknng_tls_config_from_files(
  NULL, '/tmp/ducknng_readme_http_demo_ca.pem', NULL,
  0
);

-- GET request over HTTPS.
SELECT ok, status, error, body_text
FROM ducknng_ncurl(
  'https://127.0.0.1:18443/hello',
  NULL, NULL, NULL, 2000,
  getvariable('tls_http')::UBIGINT
);

-- POST with raw bytes; inspect echoed body and response headers.
SET VARIABLE echo_hdrs = '[{"name":"Content-Type","value":"application/octet-stream"}]';

SELECT ok, status, hex(body) AS body_hex,
       position('X-Test' IN headers_json) > 0 AS has_x_test
FROM ducknng_ncurl(
  'https://127.0.0.1:18443/echo',
  'POST',
  getvariable('echo_hdrs'),
  from_hex('01020304'),
  2000,
  getvariable('tls_http')::UBIGINT
);

SELECT ducknng_drop_tls_config(getvariable('tls_http')::UBIGINT);
```

```{r http-demo-cleanup, include=FALSE}
pidfile <- "/tmp/ducknng_readme_http_demo.pid"
if (file.exists(pidfile)) {
  pid <- suppressWarnings(as.integer(readLines(pidfile, warn = FALSE)))
  if (!is.na(pid)) tools::pskill(pid)
  unlink(pidfile)
}
unlink(c(script, logfile, cafile))
```

`ducknng_ncurl_aio(...)` is the unary async form and still completes with the full response body. The `ducknng_ncurl_stream_open_aio(...)` / `ducknng_ncurl_stream_recv_aio(...)` family below is the incremental form; `ducknng_ncurl_table(...)` parses a complete response body through the codec layer.

### `ducknng_start_server` on `http://` and `https://`

Pointing `ducknng_start_server(...)` at an `http://` or `https://` URL mounts the framed RPC surface at the URL path. The higher-level synchronous request, RPC, and session helpers switch carriers automatically from the scheme; `contexts = 1` is required for the HTTP carrier.

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'http_rpc', 'http://127.0.0.1:18444/_ducknng', 1, 134217728, 300000, 0::UBIGINT
);

WITH m AS (
  SELECT manifest
  FROM ducknng_get_rpc_manifest('http://127.0.0.1:18444/_ducknng', 0::UBIGINT)
  WHERE ok
)
SELECT json_extract_string(manifest::JSON, '$.server.name') AS server_name,
       json_array_length(json_extract(manifest::JSON, '$.methods')) AS method_count
FROM m;

-- ducknng_ncurl can call the framed RPC path directly too.
SET VARIABLE http_frame = (
  SELECT body FROM ducknng_ncurl(
    'http://127.0.0.1:18444/_ducknng', 'POST',
    '[{"name":"Content-Type","value":"application/vnd.ducknng.frame"}]',
    from_hex('01000000000000000000000000000000000000000000'),
    2000, 0::UBIGINT
  )
);
SELECT ok, type_name, name FROM ducknng_decode_frame(getvariable('http_frame'));

SELECT ducknng_stop_server('http_rpc');
```

### Custom HTTP routes

HTTP and HTTPS services can register additional exact, prefix, or template routes beside the framed RPC mount. Route handlers are ordinary SQL queries that return one response row and can inspect the in-flight request through `ducknng_http_request()` and `ducknng_http_request_body()`.

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'http_routes', 'http://127.0.0.1:18445/_ducknng', 1, 134217728, 300000, 0::UBIGINT
);

SELECT ducknng_register_http_route(
  'http_routes', 'GET', '/healthz',
  'SELECT * FROM ducknng_http_text(200, ''ok'')'
);

SELECT ducknng_register_http_route(
  'http_routes', 'POST', '/echo',
  'SELECT * FROM ducknng_http_text(
     201,
     (SELECT method || '' '' || path || '' x='' ||
             coalesce(ducknng_http_query_param(''x''), '''') ||
             '' body='' || coalesce(body_text, '''')
      FROM ducknng_http_request(), ducknng_http_request_body()))'
);

SELECT ducknng_register_http_route_pattern(
  'http_routes', 'GET', 'template',
  '/tenant/{tenant_id}/items/{item_id}',
  'SELECT * FROM ducknng_http_text(
     200,
     (SELECT ducknng_http_path_param(''tenant_id'') || '':'' ||
             ducknng_http_path_param(''item_id'')
      FROM ducknng_http_request()))'
);
```

```{duckdb collapse=TRUE, comment=""}
SELECT ok, status, body_text
FROM ducknng_ncurl('http://127.0.0.1:18445/healthz', 'GET', NULL, NULL, 2000, 0::UBIGINT);

SELECT ok, status, body_text
FROM ducknng_ncurl(
  'http://127.0.0.1:18445/echo?x=1', 'POST',
  '[{"name":"Content-Type","value":"text/plain"}]',
  'hello'::BLOB, 2000, 0::UBIGINT
);

SELECT ok, status, body_text
FROM ducknng_ncurl(
  'http://127.0.0.1:18445/tenant/alice/items/42', 'GET', NULL, NULL, 2000, 0::UBIGINT
);

SELECT service_name, method, match_kind, path
FROM ducknng_list_http_routes()
WHERE service_name = 'http_routes'
ORDER BY match_kind, path;

SELECT ducknng_stop_server('http_routes');
```

Route handlers run on the `shared_serialized_connection` lane by default. Switch to `service_serialized_connection` or `request_connection` with `ducknng_set_service_execution_model(...)` before traffic arrives. The subscriber gateway walkthrough in `demo/subscriber_gateway.Rmd` (and `docs/subscriber_gateway_demo.md`) shows all three services sharing one DuckDB runtime without deadlocking because each uses `service_serialized_connection`.

The runtime execution-connection pool opens a warm minimum and grows on demand up to a deterministic ceiling. Use the SQL controls below when a workload intentionally holds many query sessions or nested `query_rpc` scans at once.

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_execution_pool_max() AS current_pool_max;
SELECT ducknng_set_execution_pool_max(96) AS effective_pool_max;
SELECT ducknng_execution_pool_max() AS updated_pool_max;
SELECT ducknng_set_execution_pool_max(64) AS reset_pool_max;
```

### Chunked streaming routes (SSE)

`ducknng_add_stream_route` registers a route that responds with HTTP chunked transfer encoding. The SQL query must return a `chunk` column; each non-null row is written to the client as a separate chunk before the terminator. `ducknng_format_sse(data, event, id, retry)` formats a single Server-Sent Events event string.

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server(
  'http_sse', 'http://127.0.0.1:18446/_ducknng', 1, 134217728, 300000, 0::UBIGINT
);

SELECT ducknng_add_stream_route(
  'http_sse', 'GET', '/events',
  'SELECT ducknng_format_sse(''row '' || i::VARCHAR) AS chunk
   FROM generate_series(1, 3) t(i)'
);

SELECT is_stream, stream_content_type
FROM ducknng_list_http_routes()
WHERE service_name = 'http_sse' AND path = '/events';
```

`ducknng_ncurl` transparently reassembles the chunked response. The body is the concatenation of all SSE events written by the server.

```{duckdb collapse=TRUE, comment=""}
SELECT ok, status, headers_json, body_text
FROM ducknng_ncurl('http://127.0.0.1:18446/events', 'GET', NULL, NULL, 2000, 0::UBIGINT);
```

For incremental consumption, the open aio completes as soon as the response status and headers arrive. Each receive aio then returns raw application body bytes with HTTP chunk framing removed. Completion is a separate `end_of_stream = true` row, and the claimed stream is closed independently of its aio handles.

```{duckdb collapse=TRUE, comment=""}
SET VARIABLE sse_open_aio = ducknng_ncurl_stream_open_aio(
  'http://127.0.0.1:18446/events', 'GET', NULL, NULL, 2000, 0::UBIGINT
);
CREATE TEMP TABLE sse_open AS
SELECT * FROM ducknng_ncurl_stream_open_aio_collect(
  list_value(getvariable('sse_open_aio')::UBIGINT), 2000
);
SET VARIABLE sse_stream = (SELECT stream_id FROM sse_open WHERE ok);

SET VARIABLE sse_recv_1 = ducknng_ncurl_stream_recv_aio(
  getvariable('sse_stream')::UBIGINT, 65536::UBIGINT, 2000
);
SELECT replace(decode(body), chr(10), '\\n') AS event, end_of_stream
FROM ducknng_ncurl_stream_recv_aio_collect(
  list_value(getvariable('sse_recv_1')::UBIGINT), 2000
);

-- Further receives return the remaining events and finally explicit EOF.
SET VARIABLE sse_recv_2 = ducknng_ncurl_stream_recv_aio(
  getvariable('sse_stream')::UBIGINT, 65536::UBIGINT, 2000
);
SET VARIABLE sse_body_2 = (
  SELECT body FROM ducknng_ncurl_stream_recv_aio_collect(
    list_value(getvariable('sse_recv_2')::UBIGINT), 2000
  )
);
SET VARIABLE sse_recv_3 = ducknng_ncurl_stream_recv_aio(
  getvariable('sse_stream')::UBIGINT, 65536::UBIGINT, 2000
);
SET VARIABLE sse_body_3 = (
  SELECT body FROM ducknng_ncurl_stream_recv_aio_collect(
    list_value(getvariable('sse_recv_3')::UBIGINT), 2000
  )
);
SET VARIABLE sse_recv_eof = ducknng_ncurl_stream_recv_aio(
  getvariable('sse_stream')::UBIGINT, 65536::UBIGINT, 2000
);
SELECT body, end_of_stream
FROM ducknng_ncurl_stream_recv_aio_collect(
  list_value(getvariable('sse_recv_eof')::UBIGINT), 2000
);

SELECT ducknng_ncurl_stream_close(getvariable('sse_stream')::UBIGINT);
SELECT ducknng_aio_drop(getvariable('sse_open_aio')::UBIGINT)
   AND ducknng_aio_drop(getvariable('sse_recv_1')::UBIGINT)
   AND ducknng_aio_drop(getvariable('sse_recv_2')::UBIGINT)
   AND ducknng_aio_drop(getvariable('sse_recv_3')::UBIGINT)
   AND ducknng_aio_drop(getvariable('sse_recv_eof')::UBIGINT) AS dropped;
DROP TABLE sse_open;

SELECT ducknng_stop_server('http_sse');
```

## Client interop: nanonext REQ/REP

This example shows a `nanonext` R client talking to a `ducknng` server using the versioned wire envelope directly. It covers frame encoding and decoding at the R level so the wire contract is explicit.

### Frame helpers

```{r exec-example-frame-helpers}
write_u32le <- function(x) writeBin(as.integer(x), raw(), size = 4L, endian = "little")
write_u64le <- function(x) {
  x <- as.double(x)
  c(write_u32le(x %% 2^32), write_u32le(floor(x / 2^32)))
}
read_u32le_bytes <- function(x) sum(as.double(as.integer(x)) * 256^(0:3))
read_u64le_bytes <- function(x) read_u32le_bytes(x[1:4]) + 2^32 * read_u32le_bytes(x[5:8])
read_u32le <- function(buf, offset) read_u32le_bytes(buf[offset + 0:3])
read_u64le <- function(buf, offset) read_u64le_bytes(buf[offset + 0:7])

encode_ducknng_exec_payload <- function(sql, want_result = FALSE) {
  raw_con <- rawConnection(raw(), open = "r+")
  on.exit(close(raw_con), add = TRUE)
  nanoarrow::write_nanoarrow(data.frame(sql = sql, want_result = want_result), raw_con)
  rawConnectionValue(raw_con)
}

encode_ducknng_call_frame <- function(method, payload, flags = 0L) {
  m <- charToRaw(method)
  c(as.raw(1L), as.raw(1L), write_u32le(flags),
    write_u32le(length(m)), write_u32le(0),
    write_u64le(length(payload)), m, payload)
}

encode_ducknng_exec_request <- function(sql, want_result = FALSE) {
  encode_ducknng_call_frame("exec", encode_ducknng_exec_payload(sql, want_result))
}

decode_ducknng_exec_reply <- function(buf) {
  if (!is.raw(buf) || length(buf) < 22L)
    stop("ducknng reply frame was empty or truncated", call. = FALSE)
  name_len    <- read_u32le(buf, 7)
  error_len   <- read_u32le(buf, 11)
  payload_len <- read_u64le(buf, 15)
  ns <- 23L; es <- ns + name_len; ps <- es + error_len
  payload <- if (payload_len > 0) buf[ps:(ps + payload_len - 1L)] else raw()
  list(
    version = as.integer(buf[1]), type = as.integer(buf[2]),
    method  = rawToChar(buf[ns:(es - 1L)]),
    error   = if (error_len > 0) rawToChar(buf[es:(ps - 1L)]) else "",
    data    = if (payload_len > 0) as.data.frame(nanoarrow::read_nanoarrow(payload)) else NULL
  )
}
```

### Start a server, dial a REQ socket, run remote SQL

```{r exec-example-setup}
ext_path <- normalizePath("build/release/ducknng.duckdb_extension")
ipc_path <- "/tmp/ducknng_readme_exec.ipc"
ipc_url  <- paste0("ipc://", ipc_path)
unlink(ipc_path)

db_con <- DBI::dbConnect(duckdb::duckdb(config = list(allow_unsigned_extensions = "true", allow_extensions_metadata_mismatch = "true")))
DBI::dbExecute(db_con, sprintf("LOAD '%s'", ext_path))
DBI::dbGetQuery(db_con, sprintf(
  "SELECT ducknng_start_server('sql_exec', '%s', 1, 134217728, 300000, 0::UBIGINT)", ipc_url
))
DBI::dbGetQuery(db_con, "SELECT ducknng_register_exec_method()")

deadline <- Sys.time() + 5
while (!file.exists(ipc_path) && Sys.time() < deadline) Sys.sleep(0.1)
req <- nanonext::socket("req", dial = ipc_url)
```

```{r exec-example-run}
send_recv <- function(sock, frame, timeout_ms = 1000L) {
  nanonext::send(sock, frame, mode = "raw", block = timeout_ms)
  decode_ducknng_exec_reply(nanonext::recv(sock, mode = "raw", block = timeout_ms))
}

# DDL.
send_recv(req, encode_ducknng_exec_request(
  "CREATE TABLE exec_demo(i INTEGER)"
))$data

# DML.
send_recv(req, encode_ducknng_exec_request(
  "INSERT INTO exec_demo VALUES (42),(99)"
))$data

# Query with want_result = TRUE returns Arrow-decoded rows.
send_recv(req, encode_ducknng_exec_request(
  "SELECT i, i > 50 AS gt_50 FROM exec_demo ORDER BY i",
  want_result = TRUE
))$data
```

The same table is visible locally through the DuckDB connection that hosts the server:

```{r exec-example-local}
DBI::dbGetQuery(db_con, "SELECT * FROM exec_demo ORDER BY i")
```

```{r exec-example-cleanup}
DBI::dbGetQuery(db_con, "SELECT ducknng_stop_server('sql_exec')")
close(req)
DBI::dbDisconnect(db_con)
unlink(ipc_path)
```

## Deployment and admission policy

`ducknng` defaults to default-open transport behavior. The application or deployment decides where the trust boundary is. For anything beyond process-local, filesystem-local, or loopback-only development, set the boundary deliberately before exposing a service.

A strong direct-exposure baseline:

1. Use `https://`, `wss://`, or `tls+tcp://` rather than plaintext network carriers.
2. Create the TLS handle with `auth_mode = 2` so peers must present a verified client certificate.
3. Configure a private CA or otherwise narrow certificate issuance so the trust root is not broader than the application.
4. Set exact peer-identity and/or IP/CIDR allowlists before exposure with `ducknng_set_tls_peer_allowlist(...)` and the optional `ip_allowlist_json` startup argument.
5. Set basic service resource limits with `ducknng_set_service_limits(name, max_open_sessions[, ...])` before exposing query sessions.
6. Install a service-level SQL authorizer with `ducknng_set_service_authorizer(...)` when admission depends on deployment tables, tenants, method names, or HTTP metadata.
7. Rotate policy dynamically with `ducknng_set_service_peer_allowlist(...)`, `ducknng_set_service_ip_allowlist(...)`, `ducknng_set_service_limits(...)`, and `ducknng_set_service_authorizer(...)`.

`exec` and query sessions run DuckDB SQL as the service process. Do not expose them to callers who are not already authorized to run that SQL. Profiles by use case:

- **Local development / single-user IPC**: `inproc://`, loopback, or `ipc://` under a private directory; `exec` may be enabled for convenience.
- **Trusted service mesh**: `tls+tcp://`, `wss://`, or `https://` with mTLS, exact peer allowlists, IP/CIDR allowlists, and service limits.
- **Shared or semi-trusted clients**: authorize both `query_open` and `exec`; query sessions are streaming, not a read-only policy. Use deployment-level DuckDB/process controls for filesystem, extension loading, outbound network, and attachments.
- **Public/untrusted internet**: do not expose raw DuckDB SQL directly. Put `ducknng` behind a gateway that authenticates, rate-limits, and maps users to fixed application operations.

```sql
-- Set a peer allowlist on the TLS handle before starting services with it.
SELECT ducknng_set_tls_peer_allowlist(
  1::UBIGINT,
  '["tls:san:client-a.example","tls:san:client-b.example"]'
);

-- Start with an IP/CIDR allowlist installed before the listener accepts traffic.
SELECT ducknng_start_server(
  'sql_http', 'http://127.0.0.1:18444/_ducknng',
  1, 134217728, 300000, 0::UBIGINT,
  '["127.0.0.1/32"]'
);

-- Set a basic resource limit for query sessions.
SELECT ducknng_set_service_limits('sql_http', 16::UBIGINT);

-- Add a flexible SQL callback for policy that depends on request context.
SELECT ducknng_set_service_authorizer(
  'sql_http',
  'SELECT remote_ip = ''127.0.0.1'' AND rpc_method = ''manifest'' AS allow,
          403 AS status, ''not authorized'' AS reason
   FROM ducknng_auth_context()'
);
```

Transport/carrier scheme reference:

| Surface | Accepted schemes | TLS handle accepted on |
| --- | --- | --- |
| `ducknng_start_server(...)` | `inproc://`, `ipc://`, `tcp://`, `tls+tcp://`, `ws://`, `wss://`, `http://`, `https://` | `tls+tcp://`, `wss://`, `https://` |
| synchronous RPC/session helpers | NNG schemes + `http://`, `https://` | `tls+tcp://`, `wss://`, `https://` |
| raw RPC AIO helpers | NNG schemes + `http://`, `https://` | `tls+tcp://`, `wss://`, `https://` |
| generic socket APIs | `inproc://`, `ipc://`, `tcp://`, `tls+tcp://`, `ws://`, `wss://` | `tls+tcp://`, `wss://` |
| HTTP client helpers | `http://`, `https://` | `https://` |

Supplying a non-zero TLS handle on a non-TLS URL is rejected rather than silently ignored.

For deeper write-ups see `docs/protocol.md`, `docs/manifest.md`, `docs/security.md`, `docs/registry.md`, `docs/transports.md`, and `docs/http.md`. `NEWS.md` summarizes landed changes.

## Introspection examples

### DuckDB log entries

`ducknng_log_entries()` returns a snapshot of the most recent DuckDB log entries captured by the extension's in-memory ring buffer (capacity 512). The ring is populated via the DuckDB logger API. Log capture is not active by default; call `ducknng_enable_log_capture()` once to register the ring with DuckDB's logger before querying the ring. Note: `ducknng_enable_log_capture()` is unsafe to call under DuckDB v1.5.2 due to a crash in the logger registration API; the example below is shown but not executed.

```{duckdb collapse=TRUE, comment="", eval=FALSE}
SELECT ducknng_enable_log_capture();

SELECT ts, level, log_type, LEFT(message, 80) AS message
FROM ducknng_log_entries()
ORDER BY ts
LIMIT 5;
```

### NNG statistics

`ducknng_nng_stats()` exposes NNG's process-wide statistics tree as rows. It is a low-level observability primitive: filter by statistic name, unit, or scope to inspect socket counters without making NNG types part of the public RPC method surface.

```{duckdb include=FALSE}
SELECT ducknng_start_server('stats_demo', 'tcp://127.0.0.1:0', 1, 134217728, 30000, 0::UBIGINT);
```

```{duckdb collapse=TRUE, comment=""}
SELECT name, unit, count(*) AS rows
FROM ducknng_nng_stats()
WHERE name IN ('tx_bytes', 'rx_bytes', 'tx_msgs', 'rx_msgs')
GROUP BY name, unit
ORDER BY name, unit;
```

```{duckdb include=FALSE}
SELECT ducknng_stop_server('stats_demo');
```

### Pipe-event monitor

Each running service maintains a bounded ring of NNG pipe events (ADD_PRE, ADD_POST, REM_POST). `ducknng_read_monitor(name, after_seq, max_events)` pages through the ring. `ducknng_monitor_status(name)` returns ring and active-pipe counters.

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_start_server('monitor_demo', 'tcp://127.0.0.1:0', 1, 134217728, 30000, 0::UBIGINT);
SET VARIABLE monitor_url = (SELECT listen FROM ducknng_list_servers() WHERE name = 'monitor_demo');
```

```{duckdb collapse=TRUE, comment=""}
-- Dial a REQ socket to generate pipe-connect events.
SET VARIABLE monitor_req = (ducknng_open_socket('req')).socket_id;
SELECT (ducknng_dial_socket(getvariable('monitor_req')::UBIGINT, getvariable('monitor_url'), 0, 0::UBIGINT)).ok AS dialed;
```

```{duckdb collapse=TRUE, comment=""}
SELECT event_capacity, event_count, oldest_seq, newest_seq,
       dropped_events, active_pipes, max_active_pipes
FROM ducknng_monitor_status('monitor_demo');
```

`ducknng_list_pipes(name)` lists currently open NNG pipes for a service.

```{duckdb collapse=TRUE, comment=""}
SELECT count(*) AS active_pipes,
       bool_and(remote_addr IS NOT NULL) AS addresses_present,
       bool_and(peer_identity IS NULL) AS anonymous
FROM ducknng_list_pipes('monitor_demo');
```

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_close_socket(getvariable('monitor_req')::UBIGINT);
SELECT ducknng_stop_server('monitor_demo');
```

Client sockets can also opt into their own pipe-event ring. Enable monitoring before the socket listens or dials if you need to capture the first add event.

```{duckdb include=FALSE}
SET VARIABLE socket_mon_a = (ducknng_open_socket('pair')).socket_id;
```

```{duckdb collapse=TRUE, comment=""}
SELECT ducknng_monitor_socket(getvariable('socket_mon_a')::UBIGINT) AS monitoring_enabled;
```

```{duckdb include=FALSE}
SELECT (ducknng_listen_socket(getvariable('socket_mon_a')::UBIGINT,
                              'inproc://ducknng_readme_socket_monitor',
                              1048576,
                              0::UBIGINT)).ok;
SET VARIABLE socket_mon_b = (ducknng_open_socket('pair')).socket_id;
SELECT (ducknng_dial_socket(getvariable('socket_mon_b')::UBIGINT,
                            'inproc://ducknng_readme_socket_monitor',
                            1000,
                            0::UBIGINT)).ok;
```

```{duckdb collapse=TRUE, comment=""}
SELECT event, count(*) AS events
FROM ducknng_read_socket_monitor(getvariable('socket_mon_a')::UBIGINT,
                                 0::UBIGINT,
                                 0::UBIGINT)
GROUP BY event
ORDER BY event;

SELECT ducknng_socket_monitor_wait(getvariable('socket_mon_a')::UBIGINT,
                                   0::UBIGINT,
                                   1000::UBIGINT) >= 1 AS saw_event;
```

```{duckdb include=FALSE}
SELECT (ducknng_close_socket(getvariable('socket_mon_a')::UBIGINT)).ok;
SELECT (ducknng_close_socket(getvariable('socket_mon_b')::UBIGINT)).ok;
```

## Browser and WebAssembly

ducknng builds as a duckdb-wasm side module, but a browser sandbox cannot offer raw sockets, so support follows the compiled capability contract rather than native assumptions. The release-supported `wasm_eh` lane covers extension load and scalar SQL, synchronous browser HTTP(S), real asynchronous unary Fetch with timeout and cancellation, body parsing, HTTPS/CORS, framed HTTP RPC/session routing, and the async ducknng-frame-over-WebSocket carrier. Incremental HTTP response streams, raw `tcp://`, `ipc://`, and native POSIX-style `tls+tcp://` remain unsupported in the browser.

The browser checks are not run during README rendering because they require Docker, a staged duckdb-wasm site, and real Chromium. They are still literate, executable chunks; run them from the repository root when validating a wasm/browser change.

```{bash wasm-eh-stage, eval=FALSE}
DUCKDB_WASM_PLATFORM=wasm_eh DUCKNNG_WASM_SERVE=0 \
  scripts/start_duckdb_wasm_local_test.sh
```

```{bash wasm-eh-smoke, eval=FALSE}
node test/browser/run_smoke.mjs .duckdb-wasm-local-artifacts/site \
  --probes=conformance
```

The threaded runtime remains diagnostic rather than release-blocking. It can build and individual HTTP probes can pass, but repeated headless runs still expose extension-load, `inproc://` progress, and DuckDB-wasm JSON-autoload instability.

```{bash wasm-threads-diagnostic, eval=FALSE}
DUCKDB_WASM_PLATFORM=wasm_threads DUCKNNG_WASM_SERVE=0 \
  scripts/start_duckdb_wasm_local_test.sh
node test/browser/run_smoke.mjs .duckdb-wasm-local-artifacts/site \
  --probes=load
```

Self-published unsigned release assets are built by the dedicated release workflow. Push release tags one at a time using `v<ducknng-version>-duckdb<duckdb-version>` so GitHub runs the binary matrix for each pinned DuckDB runtime.

```{bash release-binary-workflow, eval=FALSE}
gh workflow run ducknng-release-binaries.yml --ref main
gh run list --workflow "ducknng release binaries" --limit 5
```

The browser transport matrix, the COOP/COEP requirements for the threaded runtime, and the smoke harness are tracked in `docs/wasm.md`, `docs/wasm_browser_transport_checklist.md`, and `test/browser/README.md`.

## Status

Version 0.1.2.9000. Every DuckDB chunk shown as runnable in this README is implemented, tested, and runs at render time. Browser/WebAssembly and release-matrix checks require external runtimes, so their exact commands are shown above and covered by the Playwright and GitHub Actions harnesses. The extension builds against DuckDB v1.5.2 using the C API only.

## References

- [NNG](https://nng.nanomsg.org/) — underlying messaging library and transport family.
- [`nanonext`](https://github.com/r-lib/nanonext) — main client/server ergonomics reference.
- [`mangoro`](https://github.com/sounkou-bioinfo/mangoro) — thin-envelope + Arrow IPC RPC client.
- [`quack`](https://github.com/duckdb/duckdb-quack) — DuckDB's experimental Quack HTTP protocol extension.
- [`@quack-protocol/sdk`](https://github.com/tobilg/quack-protocol) and [`adbc-driver-quack`](https://github.com/gizmodata/adbc-driver-quack) — related Quack client/codec implementations used as protocol references.
- [DuckDB C API](https://duckdb.org/docs/stable/clients/c/api) — extension and SQL integration boundary.
- [Apache Arrow IPC](https://arrow.apache.org/docs/format/Columnar.html#serialization-and-interprocess-communication-ipc) — tabular payload format.
