---
engine: knitr
format: gfm
---

```{r}
#| include: false
piknit::register_engines()
```

# pi-ducknng

`pi-ducknng` lets Pi agents discover and call manifested ducknng endpoints.
It ships two endpoints: a persistent R session and a durable coordination
service for Pi sessions.

## Architecture

![Generic ducknng endpoint architecture](man/figures/architecture.svg)

An endpoint is placed first and supplies a URL. Discovery and calls stay
generic. DuckDB owns native extension loading and host-language calls. The
hard-vendored ducknng release owns transport, mbedTLS, identity, framing,
manifests, sessions, AIO, cancellation, and codecs. The R endpoints compose
`nanonext` and `mirai` and do not reproduce those layers. `make architecture`
regenerates the SVG from `man/figures/architecture.mmd`.

## Pi package

Install the repository as a Pi package:

```sh
pi install git:github.com/sounkou-bioinfo/pi-ducknng
```

The package contributes three model tools:

- `persistent_r_start` places the R endpoint;
- `ducknng_describe` reads any compatible endpoint manifest;
- `ducknng_call` invokes a declared method.

On first use the package builds the pinned ducknng source, which requires
Git, Make, CMake, Python, and a C/C++ toolchain. The R runtime must provide
the packages listed under `Imports` in [`DESCRIPTION`](DESCRIPTION).

## Persistent R

The evaluated cell below loads the package extension into an OpenAI Codex
agent. The model discovers the endpoint's methods and request examples, then
runs an `mtcars` analysis whose intermediate table survives across fresh
DuckDB clients.

```{pi, provider="openai-codex", model="gpt-6-astra", thinking="medium", timeout=900, no_extensions=TRUE, extension="./extensions/pi-ducknng/index.ts"}
Using the package tools, start the R adapter and follow its ducknng manifest. In one eval call, create and return `mpg_by_cyl <- aggregate(mpg ~ cyl, data = datasets::mtcars, FUN = mean)`. In a separate eval call, use the persisted `mpg_by_cyl` to return `transform(mpg_by_cyl, delta_from_4cyl = mpg - mpg[cyl == 4])`. Close the endpoint. Report the manifested methods and decoded rows, beginning with `AGENT_DUCKNNG_MANIFEST_CALL_OK` only when both calls used the same endpoint process and succeeded.
```

Each tool request opens and closes a fresh DuckDB instance. Eval results
travel as Arrow IPC streams written by nanoarrow and decoded by ducknng. The
mirai-owned R environment stays in the endpoint process until `close` or 30
seconds without a request.

The README is rendered by `piknit`, and `make readme` rejects output that
lacks the agent's success receipt.

## Durable agent coordination

Start the coordination endpoint independently of any Pi session, with its
locator and DuckDB state in a private directory:

```sh
install -d -m 700 "$HOME/.local/state/pi-ducknng"
Rscript --vanilla tools/pi-coordination-endpoint.R \
  "$HOME/.local/state/pi-ducknng/coordination.url" \
  "$HOME/.local/state/pi-ducknng/coordination.duckdb"
```

Attach an interactive Pi session in another terminal:

```sh
export PI_DUCKNNG_COORDINATION_URL="$(cat \
  "$HOME/.local/state/pi-ducknng/coordination.url")"
export PI_DUCKNNG_COORDINATION_PROJECT="my-project"
export PI_DUCKNNG_AGENT_ID="reviewer"
pi
```

The extension registers the session and polls the `reviewer` mailbox. It
injects each message through Pi's steering API and acknowledges it once that
call returns.

- `send` stores mail before replying. Queued mail and reservations survive
  disconnected sessions and endpoint restarts.
- Delivery is at least once. A crash between injection and acknowledgement
  redelivers the same message ID, and work done in response is not exactly
  once.
- Reservations are advisory leases over `resource:` identifiers or canonical
  `file:///` URIs, with a fencing value that increases across the project.
- Acknowledged messages are kept for 30 days. A mailbox holds at most 10,000
  unacknowledged messages.

This first endpoint is for one trusted user on one machine. It listens on a
new loopback port at each start, so restart attached Pi sessions after
restarting it. It has no authentication, and the adapter does not yet give
the model its own send or reserve tool. NNG PUB/SUB is kept off the
correctness path: a later event socket may reduce wake-up latency, but
`receive` always repairs missed events.

## Agent-backed pkgdown articles

Two pkgdown articles are authored as `vignettes/*.Rmd.orig` and precomputed
with `openai-codex` agents. The committed `.Rmd` results keep package and site
builds credential-free, following the
[rOpenSci precomputed-vignette pattern](https://ropensci.org/blog/2019/12/08/precompute-vignettes/).

```sh
make vignettes
make site
```

## Pinned runtime tuple

```{r}
#| echo: false
pins <- read.table(
  "DEPENDENCIES", sep = "=", quote = "", comment.char = "",
  col.names = c("key", "value")
)
pin <- function(key) pins$value[pins$key == key]
knitr::kable(data.frame(
  Component = c(
    "DuckDB", "`@duckdb/node-api`", "ducknng",
    "`@earendil-works/pi-coding-agent`", "`@earendil-works/pi-agent-core`"
  ),
  Version = c(
    pin("DUCKDB_VERSION"), pin("DUCKDB_NODE_API_VERSION"),
    sprintf("`%s`", pin("DUCKNNG_REF")),
    pin("PI_CODING_AGENT_VERSION"), pin("PI_AGENT_CORE_VERSION")
  )
))
```

[`DEPENDENCIES`](DEPENDENCIES) is the version authority and records the exact
source commits.
