---
engine: knitr
format: gfm
---

```{r}
#| include: false
piknit::register_engines()
```

# pi-ducknng

<a href="https://lifecycle.r-lib.org/articles/stages.html#experimental"><img src="https://img.shields.io/badge/lifecycle-experimental-orange.svg" alt="Lifecycle: experimental" /></a>

`pi-ducknng` lets Pi agents discover and call manifested ducknng endpoints.
It ships two endpoints: a persistent R session and a durable coordination
service for Pi sessions.

## Architecture

![Generic ducknng endpoint architecture](man/figures/architecture.svg)

An endpoint is placed first and supplies a URL. Discovery and calls stay
generic. DuckDB owns native extension loading and host-language calls. The
hard-vendored ducknng release owns transport, mbedTLS, identity, framing,
manifests, sessions, AIO, cancellation, and codecs. The R endpoint composes
`nanonext` and `mirai` and does not reproduce those layers. The coordination
endpoint has no host-language logic at all: its methods are SQL that ducknng
serves. `make architecture` regenerates the SVG from
`man/figures/architecture.mmd`.

## Pi package

Install the repository as a Pi package:

```sh
pi install git:github.com/sounkou-bioinfo/pi-ducknng
```

The package contributes three generic model tools:

- `persistent_r_start` places the R endpoint;
- `ducknng_describe` reads any compatible endpoint manifest;
- `ducknng_call` invokes a declared method.

A session attached to a coordination endpoint also gets the coordination
tools described below.

On first use the package builds the pinned ducknng source, which requires
Git, Make, CMake, Python, and a C/C++ toolchain. The R runtime must provide
the packages listed under `Imports` in [`DESCRIPTION`](DESCRIPTION).

For `tls+tcp://` and `wss://` endpoints, point `PI_DUCKNNG_TLS_CA_FILE` at the
CA that signed the server certificate. For mutual TLS, also point
`PI_DUCKNNG_TLS_CERT_KEY_FILE` at a PEM file holding the client certificate
and key. The paths never enter tool schemas or results.

## Persistent R

The evaluated cell below loads the package extension into an OpenAI Codex
agent. The model discovers the endpoint's methods and request examples, then
runs an `mtcars` analysis whose intermediate table survives across fresh
DuckDB clients.

```{pi, provider="openai-codex", model="gpt-6-astra", thinking="medium", timeout=900, no_extensions=TRUE, extension="./extensions/pi-ducknng/index.ts"}
Using the package tools, start the R adapter and follow its ducknng manifest. In one eval call, create and return `mpg_by_cyl <- aggregate(mpg ~ cyl, data = datasets::mtcars, FUN = mean)`. In a separate eval call, use the persisted `mpg_by_cyl` to return `transform(mpg_by_cyl, delta_from_4cyl = mpg - mpg[cyl == 4])`. Close the endpoint. Report the manifested methods and decoded rows, beginning with `AGENT_DUCKNNG_MANIFEST_CALL_OK` only when both calls used the same endpoint process and succeeded.
```

Each tool request opens and closes a fresh DuckDB instance. Eval results
travel as Arrow IPC streams written by nanoarrow and decoded by ducknng. The
mirai-owned R environment stays in the endpoint process until `close` or
until the Pi process that started it exits.

The README is rendered by `piknit` with the Pi version pinned in
`DEPENDENCIES`. `make readme` rejects output that lacks any agent's success
receipt.

## Durable agent coordination

The coordination endpoint is a DuckDB database whose methods are
[SQL files](coordination/methods) served by ducknng. Start it independently of
any Pi session, with its state in a private directory. It listens on `ipc://`
beside the database, so the address survives restarts and only the
directory's owner can connect:

```sh
install -d -m 700 "$HOME/.local/state/pi-ducknng"
node tools/pi-coordination-endpoint.ts \
  "$HOME/.local/state/pi-ducknng/coordination.url" \
  "$HOME/.local/state/pi-ducknng/coordination.duckdb"
```

Attach a Pi session in another terminal:

```sh
export PI_DUCKNNG_COORDINATION_URL="$(cat \
  "$HOME/.local/state/pi-ducknng/coordination.url")"
export PI_DUCKNNG_COORDINATION_PROJECT="my-project"
export PI_DUCKNNG_AGENT_ID="reviewer"
pi
```

The attached session gets five tools: `coordination_send`,
`coordination_inbox`, `coordination_agents`, `coordination_reserve`, and
`coordination_release`. The registration behind them never enters the
model's context.

- **Delivery.** An interactive session polls its mailbox in the background
  and steers each message into the conversation. A one-shot `pi -p` run pulls
  mail with `coordination_inbox`. Every message arrives framed with its sender
  and message ID and is marked as coming from another agent, not the user.
- **Durability.** `send` stores mail before replying, so mail survives
  disconnected sessions and endpoint restarts. Delivery is at least once, and
  work done in response is not exactly once. Mail nobody receives becomes a
  dead letter after 7 days, and `send` reports recipients that have never
  registered.
- **Reservations.** Leases cover `resource:` identifiers or paths and carry a
  fencing value that increases across the project. The session renews each
  lease it holds until release or shutdown. While another session holds a
  path, Pi's `edit` and `write` tools refuse to touch it.

Over `ipc://`, filesystem permissions are the only authentication. To serve
agents on several hosts or users, start the endpoint on mutual TLS with a
grants file that names which certificate may act as which agent:

```sh
cat > grants.json <<'JSON'
[{"peer_identity": "tls:cn:reviewer", "project_id": "my-project", "agent_id": "reviewer"},
 {"peer_identity": "tls:cn:planner", "project_id": "my-project", "agent_id": "*"}]
JSON
PI_DUCKNNG_COORDINATION_TLS_CERT_KEY_FILE=server.pem \
PI_DUCKNNG_COORDINATION_TLS_CA_FILE=ca.pem \
PI_DUCKNNG_COORDINATION_GRANTS_FILE=grants.json \
  node tools/pi-coordination-endpoint.ts coordination.url coordination.duckdb \
  tls+tcp://0.0.0.0:7400
```

Every call must then present a certificate from that CA. A certificate may
register only the agents it is granted, and a registration ID is useless to
any other certificate.

A host process that owns a `pi-agent-core` `AgentHarness` can attach a lane
to a mailbox. Each message is committed to the lane before it is
acknowledged, and a redelivered message is matched to its existing lane entry
by message ID:

```js
import { attachCoordinationLane } from "pi-ducknng/extensions/pi-ducknng/harness.ts";
import { ducknngCoordinationClient } from "pi-ducknng/extensions/pi-ducknng/index.ts";

const coordination = attachCoordinationLane({
  client: ducknngCoordinationClient,
  url, projectId: "my-project", agentId: "worker", instanceId: "host-1",
  lane: await harness.lane("main", context),
  context,
});
await coordination.ready;
```

### Two agents, one handoff

The cells below start a coordination endpoint and run two one-shot Codex
agents in sequence. The planner mails a handoff to a reviewer that has never
registered. The reviewer pulls the handoff, answers it with persistent R,
and replies. The last cell checks the round trip through the endpoint's own
methods rather than trusting either agent's report.

```{r}
#| echo: false
state <- tempfile("pi-ducknng-readme-")
dir.create(state, mode = "0700")
locator <- file.path(state, "coordination.url")
endpoint <- processx::process$new(
  "node",
  c("tools/pi-coordination-endpoint.ts", locator, file.path(state, "coordination.duckdb"))
)
listening <- function() {
  file.exists(locator) && length(readLines(locator, warn = FALSE)) == 1L
}
while (!listening()) {
  if (!endpoint$is_alive()) stop("coordination endpoint exited")
  Sys.sleep(0.1)
}
Sys.setenv(
  PI_DUCKNNG_COORDINATION_URL = readLines(locator),
  PI_DUCKNNG_COORDINATION_PROJECT = "readme",
  PI_DUCKNNG_AGENT_ID = "planner"
)
```

```{pi, provider="openai-codex", model="gpt-6-astra", thinking="medium", timeout=900, no_extensions=TRUE, extension="./extensions/pi-ducknng/index.ts"}
You are agent `planner`. Use coordination_agents to see who is live, then reserve `resource:readme/mtcars-review`. Send agent `reviewer` a handoff of exactly two lines. Line 1: `Review mpg by cylinder count in datasets::mtcars.` Line 2: `Reply with one sentence naming the cylinder count with the highest mean mpg and that mean.` Then release the reservation. Report the live agents, the message ID, whether the recipient had been seen, and the fencing value, beginning with `AGENT_COORDINATION_SENT` only when the send and the release both succeeded.
```

```{r}
#| echo: false
Sys.setenv(PI_DUCKNNG_AGENT_ID = "reviewer")
```

```{pi, provider="openai-codex", model="gpt-6-astra", thinking="medium", timeout=900, no_extensions=TRUE, extension="./extensions/pi-ducknng/index.ts"}
You are agent `reviewer`. Call coordination_inbox with wait_ms 20000 to read your handoff. Answer it by computing the aggregate with the persistent R adapter (start it, follow its ducknng manifest, and close it when done). Reply to the sender with coordination_send in one sentence. Report the sender, the message ID, both handoff lines exactly as received, the computed rows, and your reply's message ID, beginning with `AGENT_COORDINATION_REPLIED` only when you received exactly one message, it came from `planner`, and your reply was stored.
```

The endpoint, not either model, confirms the round trip:

```{r}
#| comment: ""
source("tools/ducknng-rpc.R")
url <- Sys.getenv("PI_DUCKNNG_COORDINATION_URL")
verifier <- rpc_call(url, "register", list(
  project_id = "readme", agent_id = "planner",
  instance_id = "readme-verifier", operation_key = "verify"
))
inbox <- rpc_call(url, "receive", list(
  registration_id = verifier$registration_id
))$messages
reservations <- rpc_call(url, "list_reservations", list(
  registration_id = verifier$registration_id
))$reservations
stopifnot(
  length(inbox) == 1L,
  identical(inbox[[1L]]$sender_agent_id, "reviewer"),
  length(reservations) == 0L
)
invisible(rpc_call(url, "ack", list(
  registration_id = verifier$registration_id,
  receipt_token = inbox[[1L]]$receipt_token
)))
cat(
  "COORDINATION_ROUND_TRIP_VERIFIED",
  sprintf("reply from %s: %s", inbox[[1L]]$sender_agent_id, inbox[[1L]]$content),
  sep = "\n"
)
```

```{r}
#| include: false
endpoint$kill()
unlink(state, recursive = TRUE)
```

## Agent-backed pkgdown articles

Two pkgdown articles are authored as `vignettes/*.Rmd.orig` and precomputed
with `openai-codex` agents. The committed `.Rmd` results keep package and site
builds credential-free, following the
[rOpenSci precomputed-vignette pattern](https://ropensci.org/blog/2019/12/08/precompute-vignettes/).

```sh
make vignettes
make site
```

## Pinned runtime tuple

```{r}
#| echo: false
pins <- read.table(
  "DEPENDENCIES", sep = "=", quote = "", comment.char = "",
  col.names = c("key", "value")
)
pin <- function(key) pins$value[pins$key == key]
knitr::kable(data.frame(
  Component = c(
    "DuckDB", "`@duckdb/node-api`", "ducknng",
    "`@earendil-works/pi-coding-agent`", "`@earendil-works/pi-agent-core`"
  ),
  Version = c(
    pin("DUCKDB_VERSION"), pin("DUCKDB_NODE_API_VERSION"),
    sprintf("`%s`", pin("DUCKNNG_REF")),
    pin("PI_CODING_AGENT_VERSION"), pin("PI_AGENT_CORE_VERSION")
  )
))
```

[`DEPENDENCIES`](DEPENDENCIES) is the version authority and records the exact
source commits.
